Quantitative Benchmarking for Automatically Generated Conjectures

### 2017-09-22 TEBenchmarkPaper/article.tex
TEBenchmark additions and cleanup

% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
\journalname{Journal of Automated Reasoning}

@@

\title{Quantitative Benchmarking for Automatically Generated Conjectures%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
}
% \subtitle{Do you have a subtitle?\\ If so, write it here}

@@

\maketitle
%IGNORE ISABELLE THEORIES FOR NOW!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
%IGNORE CLUSTERING FOR NOW!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

%COMPARE COLTON ET AL.

% Evaluation of conjectures, sets of conjectures, systems
% what constitutes a good/desired output (in each author's terms)? This combines
% their evaluation with their (system-specific) assumptions (e.g. IsaCoSy doesn't
% generate redundant terms: non-redundancy is clearly a goal, but it's not measured
% in eval since it's assumed based on system's construction)

% Look carefully at each system's (paper's) eval section
% Make a table like tabl 1 of Colton et al

@@

% split into sub-sections, e.g. motivation, existing approaches, etc.

\begin{figure}
  \begin{equation*}
    \begin{split}
      \forall a. \text{Nil} &: \text{List} \  a \\
      \forall a. \text{Cons} &: a \rightarrow \text{List} \  a \rightarrow \text{List} \ a \\
      \text{head}(\text{Cons}(x, xs)) &= x \\
      \text{tail}(\text{Cons}(x, xs)) &= xs \\
      \text{append}(xs, \text{Nil}) &= xs \\
      \text{append}(\text{Cons}(x, xs), ys) &= \text{Cons}(x, \text{append}(xs, ys)) \\
      \text{reverse}(\text{Nil}) &= \text{Nil} \\
      \text{reverse}(\text{Cons}(x, xs)) &= \text{append}(\text{reverse}(xs), \text{Cons}(x, \text{Nil})) \\
      \text{length}(\text{Nil}) &= \text{Z} \\
      \text{length}(\text{Cons}(x, xs)) &= \text{S} (\text{length}(xs)) \\
      \text{map}(f, \text{Nil}) &= \text{Nil} \\
      \text{map}(f, \text{Cons}(x, xs)) &= \text{Cons}(f(x), \text{map}(f, xs)) \\
      \text{foldl}(f, x, \text{Nil}) &= x \\
      \text{foldl}(f, x, \text{Cons}(y, ys)) &= \text{foldl}(f, f(x, y), xs) \\
      \text{foldr}(f, \text{Nil}, y) &= y \\
      \text{foldr}(f, \text{Cons}(x, xs), y) &= f(x, \text{foldr}(f, xs, y))
    \end{split}
  \end{equation*}
  \caption{A simple theory defining a $\text{List}$ type and some associated
    operations. $\text{Z}$ and $\text{S}$ are from a Peano encoding of the natural numbers.
    Taken from \cite{Johansson.Dixon.Bundy:conjecture-generation}}
  \label{figure:list_theory}
\end{figure}

@@

\begin{equation} \label{eq:mapreduce}
  \forall f. \forall xs. \forall ys. \text{map}(f, \text{append}(xs, ys)) = \text{append}(\text{map}(f, xs), \text{map}(f, ys))
\end{equation}

This justifies a rewrite rule for splitting a single $\text{map}$ call into
multiple independent calls dealing with different sections of a list. Such rules
are useful optimisations since each call can be evaluated in parallel, leading
to the ``map/reduce'' programming paradigm.

@@

The task of discovering theorems in an arbitrary theory can be broken down into
two steps:

\begin{itemize}
\item \emph{Exploring} the theory, to find patterns suitable for posing as
  conjectures.
\item \emph{Proving} that those conjectures are theorems.
\end{itemize}

This second step is studied extensively in the field of Automated Theorem
Proving (ATP). Less attention has been paid to the first step, that of
Automated Theory Exploration (ATE).

A major challenge in ATE is choosing how to narrow down the set of generated
conjectures to those deemed ``interesting'', since this is an imprecise term
with many different interpretations. For example, all existing approaches agree
that simple tautologies are ``uninteresting'', but differ when it comes to more
complex statements.

A survey of ATE systems and their notions of ``interestingness'' is given by
Colton et al~\cite{colton2000notion}. Their survey also includes \emph{concept
  formation}, which corresponds to the automatic creation of theories, although
We ignore this as a separate concern: the ATE task assumes that a theory is
provided. We summarise those aspects relevant to conjecture formation in table
\ref{table:colton}, and extend the comparison with three more recent systems.

\begin{table}
  \begin{center}
    \begin{tabular}{ |l|l|c|c|c|c|c|c| }
      \hline
      \multirow{2}{*}{\textbf{Program}}                 &
      \multirow{2}{*}{\textbf{Conjecture Types}}        &
      \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\
      \hhline{~~------}
         & & \textbf{E} & \textbf{N} & \textbf{S} & \textbf{A} & \textbf{C} & \textbf{U} \\

      \hline
      AM          & if-and-only-if, implies, non-exists & X & X & X & X & X & X \\ \hline
      GT          & if-and-only-if, implies, non-exists & X & X & X & X & X & X \\ \hline
      Graffiti    & inequalities                        & X & X & X &   & X & X \\ \hline
      Bagai et al & non-exists                          &   & X &   & X & X &   \\ \hline
      HR          & if-and-only-if, implies, non-exists & X & X & X & X & X & X \\ \hline
      QuickSpec   & equalities                          & X & X &   &   & X &   \\ \hline
      IsaCoSy     & equalities                          &   & X &   &   & X &   \\ \hline
      IsaScheme   & equalities                          &   &   &   &   & X &   \\ \hline
    \end{tabular}
  \end{center}
  \caption{Classification of ATE systems from \cite{colton2000notion}, extended
    to those compared in \cite{claessen2013automating} (QuickSpec is the
    conjecture generation component of HipSpec). The interestingness measures
    are \textbf{E}mpirical plausibility, \textbf{N}ovelty,
    \textbf{S}urprisingness, \textbf{A}pplicability, \textbf{C}omprehensibility
    (low complexity) and \textbf{U}tility.}
  \label{table:colton}
\end{table}

@@

The criteria for ``interestingness'' in table \ref{table:colton} not only allow
the results of ATE systems to be evaluated, but they also affect the choice of
algorithm used by those systems in the first place. This coupling between
implementation detail and assessment method makes it difficult to compare
existing approaches, since the assumptions made by each evaluation methodology
may lead to incompatible results. For example, the design of IsaCoSy treats
simple substitutions of existing conjectures as uninteresting; yet this
criterion is absent from its evaluation, since the algorithm uses a constraint
solver to entirely avoid such results.

@@

\begin{itemize}
\item The quality of each conjecture, which is the ``interestingness'' discussed
  above.
\item The \emph{set} of conjectures produced, to assess things like consistency
  and redundancy.
\item Performance of the system, to find the balance struck between output
  quality and time taken.
\end{itemize}

Existing evaluation of these three systems have assessed the interestingness of
their output by operating in well-studied theories, and taking a set of existing
theorems as a \emph{ground truth} to compare against.

@@

% Better presentation, flowchart?
\begin{itemize}
\item Collect together all definitions found in TIP into one large theory,
  removing $\alpha$-equivalent duplicates.
\item For each type, provide extra functions which act as constructors and
  destructors. For example, the $\text{List}$ type would have extra constructor
  functions $\text{constructor-Nil}$ and $\text{constructor-Cons}$, and
  destructor functions $\text{destructor-head}$ and $\text{destructor-tail}$.
  This allows us to ignore constructors and destructors, and concentrate solely
  on functions.
\item Collect together all theorem statements found in TIP. Update any
  references to removed definitions, and replace references to constructors with
  the associated constructor function from above.
\item \emph{Sample} a sub-set of functions from the large theory uniformly, such
  that we choose all of the names referenced by at least one of the theorems
  (for example, theorem \ref{eq:mapreduce} references the names
  $\text{map}$ and $\text{append}$).
\item Provide this sub-set as an input to the ATE system, timing how long it
  takes to complete.
\item Perform precision/recall analysis on the resulting conjectures, against
  the ground truth of TIP theorems which only reference names in this sample.
\end{itemize}

>>>>

% please place your own definitions here and don't use \def but
% \newcommand{}{}

% For example theory of lists
\newcommand{\function}{\rightarrow}
\newcommand{\Zero}{\text{Z}}
\newcommand{\Succ}{\text{S}}
\newcommand{\List}{\text{List}}
\newcommand{\ListA}{\text{List} \  a}
\newcommand{\Nil}{\text{Nil}}
\newcommand{\Cons}{\text{Cons}}
\newcommand{\Head}{\text{head}}
\newcommand{\Tail}{\text{tail}}
\newcommand{\Append}{\text{append}}
\newcommand{\Reverse}{\text{reverse}}
\newcommand{\Length}{\text{length}}
\newcommand{\Map}{\text{map}}
\newcommand{\Foldl}{\text{foldl}}
\newcommand{\Foldr}{\text{foldr}}

% For interestingness table
\newcommand{\iE}{\textbf{E}}
\newcommand{\iN}{\textbf{N}}
\newcommand{\iS}{\textbf{S}}
\newcommand{\iA}{\textbf{A}}
\newcommand{\iC}{\textbf{C}}
\newcommand{\iU}{\textbf{U}}
\newcommand{\tIFF}{if-and-only-if}
\newcommand{\tNE}{non-exists}
\newcommand{\tIMP}{implies}
\newcommand{\tRow}[1]{#1 \\ \hline}

% Insert the name of "your journal" with
\journalname{Journal of Automated Reasoning}

@@

\title{Quantitative Benchmarking for Automatically Generated Conjectures%\thanks
% {Grants or other notes about the article that should go on the front page
% should be placed here. General acknowledgments should be placed at the end of
% the article.}
}
% \subtitle{Do you have a subtitle?\\ If so, write it here}

@@

\maketitle

% Look carefully at each system's (paper's) eval section

@@

\begin{figure}
  \begin{equation*}
    \begin{split}
      \forall a. \Nil            &: \ListA                                  \\
      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
      \Head(\Cons(x, xs))        &= x                                       \\
      \Tail(\Cons(x, xs))        &= xs                                      \\
      \Append(\Nil,         ys)  &= ys                                      \\
      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
      \Reverse(\Nil)             &= \Nil                                    \\
      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
      \Length(\Nil)              &= \Zero                                   \\
      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
      \Map(f, \Nil)              &= \Nil                                    \\
      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
      \Foldl(f, x, \Nil)         &= x                                       \\
      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), xs)                  \\
      \Foldr(f, \Nil,         y) &= y                                       \\
      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
    \end{split}
  \end{equation*}
  \caption{A simple theory defining a $\List$ type and some associated
    operations. $\Zero$ and $\Succ$ are from a Peano encoding of the
    natural numbers. Taken from
    \cite{Johansson.Dixon.Bundy:conjecture-generation}}
  \label{figure:list_theory}
\end{figure}

@@

\begin{equation} \label{eq:mapreduce}
  \forall f. \forall xs. \forall ys.
    \Map(f, \Append(xs, ys)) = \Append(\Map(f, xs), \Map(f, ys))
\end{equation}

This justifies a rewrite rule for splitting a single $\Map$ call into
multiple independent calls dealing with different sections of a list. Such rules
are useful optimisations since each call can be evaluated in parallel, leading
to the ``map/reduce'' programming paradigm.

@@

The task of discovering theorems in an arbitrary theory can be described as the
interplay of several processes:

% Is it clear?

\begin{itemize}
\item \emph{Forming} or refining the theory, to add new definitions and
  concepts.
\item \emph{Exploring} the theory, to find patterns suitable for posing as
  conjectures.
\item \emph{Proving} that those conjectures are theorems.
\end{itemize}

The latter is studied extensively in the field of Automated Theorem Proving
(ATP). Less attention has been paid to automating the first two, those of
\emph{automated theory formation} and \emph{automated theory exploration} (ATE).

We focus on ATE, where a major challenge is choosing how to narrow down the set
of generated conjectures to those deemed ``interesting'', since this is an
imprecise term with many different interpretations. For example, all existing
approaches agree that simple tautologies are ``uninteresting'', but differ when
it comes to more complex statements.

A survey of systems for theory formation and exploration, and their notions of
``interestingness'' (for concepts and conjectures), is given by Colton et
al~\cite{colton2000notion}. Six notions of ``interestingness'' are identified in
these systems, their use by each is summarised in table \ref{table:colton},
along with (our interpretation of) their use in three more recent ATE systems.
These notions, as applied to ATE, are:

\textbf{Empirical plausibility}, which checks whether a property holds across
some specific examples. This is especially useful for avoiding  false
conjectures, without resorting to a full proof search.

\textbf{Novelty} is whether a conjecture, or one isomorphic or more general, has
already been seen.

\textbf{Surprisingness} of a conjecture is whether or not it is ``obvious'', for
example if it is an instance of a tautology.

\textbf{Applicability} depends on the number of models in which a conjecture
holds. The system of Bagai et al conjectures the \emph{non-existence} of
objects, and hence favours statements with \emph{zero} applicability. Other
systems treat applicability as a positive aspect: the more applicable the
statement, the more interesting it is.

\textbf{Comprehensibility} is the \emph{complexity} of a statement. Simpler
statements are considered more interesting, and many of the search algorithms
explore simpler/smaller statements before complex/larger ones, to more
efficiently find those which are interesting.

\textbf{Utility} is the relevance or usefulness of a conjecture to the user's
particular task. For example, if we want to find optimising rewrite rules such
as equation \ref{eq:mapreduce}, then utility would include whether or not a
conjecture justifies a rewrite rule, the difference in resource usage of the
expressions involved, and how common those expressions are in real usage.

\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                     &
    \multirow{2}{*}{\textbf{Conjecture Types}}            &
    \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{            &                    & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM          & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{GT          & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti    & inequalities       &   X &   X &   X &     &   X &   X}
    \tRow{Bagai et al & \tNE               &     &   X &     &   X &   X &    }
    \tRow{HR          & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{QuickSpec   & equalities         &   X &   X &     &     &   X &    }
    \tRow{IsaCoSy     & equalities         &     &   X &     &     &   X &    }
    \tRow{IsaScheme   & equalities         &     &     &     &     &   X &    }
  \end{tabular}
  \caption{Classification of ATE systems from \cite{colton2000notion}, extended
    to those compared in \cite{claessen2013automating} (QuickSpec is the
    conjecture generation component of HipSpec). The interestingness measures
    are \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

These criteria, especially empirical plausibility and comprehensibility, are not
only used for analysing results after the fact, but instead may form a core part
of an algorithm's design decisions. For example, the designers of IsaCoSy
consider simple substitutions of existing conjectures (i.e. those with low
novelty) to be uninteresting, and hence their system includes a constraint
solving component which avoids generating such statements entirely.

With such ambiguous and varied goals, approaches and assessment criteria it is
difficult to compare ATE systems in a quantitative way, and hence to form some
measure of ``progress'' for the field.

@@

\begin{itemize}
\item The quality of each conjecture, which is the ``interestingness'' discussed
  above.
\item The \emph{set} of conjectures produced, to assess things like novelty, as
  well as consistency of the results.
\item Performance of the system, to find the balance struck between output
  quality and time taken.
\end{itemize}

% TODO: Should "generality" be one? Or is that just an implicit "don't cheat"
% assumption?

@@

% TODO: precision/recall is evaluating the set of conjectures. It assumes that
% interestingness == membership in the ground truth, and ignores efficiency
% completely

@@

% Better presentation, flowchart?
\begin{itemize}
\item Collect together all definitions found in TIP into one large theory,
  removing $\alpha$-equivalent duplicates.
\item For each type, provide extra functions which act as constructors and
  destructors. For example, the $\List$ type would have extra constructor
  functions $\text{constructor-Nil}$ and $\text{constructor-Cons}$, and
  destructor functions $\text{destructor-head}$ and $\text{destructor-tail}$.
  This allows us to ignore constructors and destructors, and concentrate solely
  on functions.
\item Collect together all theorem statements found in TIP. Update any
  references to removed definitions, and replace references to constructors with
  the associated constructor function from above.
\item \emph{Sample} a sub-set of functions from the large theory uniformly, such
  that we choose all of the names referenced by at least one of the theorems
  (for example, theorem \ref{eq:mapreduce} references the names
  $\Map$ and $\Append$).
\item Provide this sub-set as an input to the ATE system, timing how long it
  takes to complete.
\item Perform precision/recall analysis on the resulting conjectures, against
  the ground truth of TIP theorems which only reference names in this sample.
\end{itemize}
